---
title: "Data Architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Package Ecosystem

The torpverse consists of three packages:

| Package | Role | Storage |
|---------|------|---------|
| **torp** | Core analytics, data loading, ratings, models | R package |
| **torpdata** | Processed AFL data | GitHub releases (parquet) |
| **torpmodels** | Pre-trained models | GitHub releases (RDS) |

## Data Flow

```
AFL API (fitzRoy) --> torp (clean + enrich) --> torpdata (store)
                          |
                     torpmodels (models)
                          |
                     torp (predict)
```

1. **Collection**: Raw data fetched from the AFL API via [fitzRoy](https://jimmyday12.github.io/fitzRoy/)
2. **Processing**: torp cleans data, adds EP/WP/shot model predictions
3. **Storage**: Processed data uploaded to torpdata GitHub releases as parquet files
4. **Loading**: torp's `load_*()` functions download from torpdata on demand
5. **Models**: Trained models stored in torpmodels, loaded via `load_model_with_fallback()`

## Data Types

| Function | Data | Approx Size |
|----------|------|------------|
| `load_pbp()` | Play-by-play events | ~320K rows/season |
| `load_chains()` | Possession sequences | ~160K rows/season |
| `load_player_stats()` | Player game stats | ~10K rows/season |
| `load_player_game_data()` | Player credit points for TORP | ~10K rows/season |
| `load_fixtures()` | Match schedule | ~200 rows/season |
| `load_results()` | Match scores | ~200 rows/season |
| `load_teams()` | Team lineups | ~10K rows/season |
| `load_player_details()` | Player bio info | ~800 rows/season |
| `load_xg()` | Expected goals | ~200 rows/season |
| `load_torp_ratings()` | Pre-computed TORP ratings | ~113K rows total |

### Loading Examples

```{r loading}
library(torp)

# Single season
pbp <- load_pbp(2025)

# Specific rounds
pbp <- load_pbp(2025, rounds = 1:10)

# All available data (2021+)
pbp_all <- load_pbp(TRUE, rounds = TRUE)
chains_all <- load_chains(TRUE)
stats_all <- load_player_stats(TRUE)
```

## Caching

### Disk Cache

All data downloads are cached to disk. Subsequent calls for the same data load from cache instead of re-downloading.

```{r disk-cache}
# Cache location
get_disk_cache_dir()
# Typically: tools::R_user_dir("torp", "cache")

# Check cache size
get_disk_cache_size()

# Clear cache
clear_disk_cache()
```

### In-Memory Cache

Fixtures are cached in-memory within a session for fast repeated access:

```{r mem-cache}
fixtures <- load_fixtures(all = TRUE)  # downloads + caches
fixtures <- load_fixtures(all = TRUE)  # returns cached copy

# Force refresh
clear_fixture_cache()
```

### Model Cache

Loaded models are cached in memory to avoid reloading:

```{r model-cache}
# First call loads model from torpmodels
pbp_ep <- add_epv_vars(pbp)

# Subsequent calls use cached model
pbp_ep2 <- add_epv_vars(pbp2)

# Clear after updating torpmodels
clear_model_cache()
```

## torpdata Releases

Processed data is organised into GitHub release tags:

| Release Tag | Contents |
|------------|----------|
| `pbp-data` | Play-by-play files by season |
| `chains-data` | Chain files by season |
| `player_stats-data` | Player stat files by season |
| `player_game-data` | Player game credit data by season |
| `fixtures-data` | Fixture files by season |
| `results-data` | Result files by season |
| `teams-data` | Team lineup files by season |
| `player_details-data` | Player detail files by season |
| `xg-data` | Expected goals by season |
| `ratings-data` | Pre-computed TORP ratings |
| `predictions` | Match predictions |

## torpmodels Releases

| Release Tag | Contents |
|------------|----------|
| `core-models` | EP, WP, shot, and match prediction models |
| `stat-models` | 58 stat-specific GAM models for player projections |

## Automated Updates

A GitHub Action (`.github/workflows/daily-data-release.yml`) runs daily at 2:00 AM AEST to:

1. Fetch latest data from the AFL API
2. Process through torp's cleaning pipeline
3. Upload to torpdata releases

It can also be triggered manually with `force_release` or `rebuild_aggregates` options.

## See Also

- `vignette("getting-started")` -- Installation and data loading
- `vignette("model-usage")` -- EP, WP, and shot models
- `vignette("player-ratings")` -- TORP rating methodology
