---
title: "TORP Player Ratings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TORP Player Ratings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

TORP (Total Overall Rating Points) is a player rating system that evaluates AFL players based on their contributions across four skill categories. Each component measures a different aspect of on-field value using play-by-play expected points data.

The overall rating is:

```
torp = recv/(wt + prior_recv) + disp/(wt + prior_disp)
     + spoil/(wt + prior_spoil) + hitout/(wt + prior_hitout)
```

where `wt` is the sum of exponentially decayed game weights and each `prior` provides Bayesian shrinkage for players with limited data.

## Rating Components

### Disposal (torp_disp)

Measures the value a player adds through kicks and handballs, based on the change in expected points (delta EPV) from each disposal. Adjusted for game context (possessing vs defending) and penalised for bounces.

### Receiving (torp_recv)

Captures the value of receiving a disposal. When a player receives the ball, the EPV change is attributed to the receiver as well as the disposer, recognising that getting to the right position creates value.

### Spoil/Tackle (torp_spoil)

Evaluates defensive impact from spoils, tackles, and pressure acts. Defensive-half pressure acts are subtracted (they indicate the opposition has progressed the ball) while other pressure and tackle actions are rewarded.

### Hitout (torp_hitout)

Quantifies ruck contribution from hitouts, hitouts to advantage, and ruck contests. Hitouts to advantage are weighted higher than general hitouts, while ruck contests (where no hitout is won) are penalised slightly.

## Getting Ratings

### Current Ratings

```{r current}
library(torp)

# Ratings for the upcoming round (uses all historical data)
ratings <- calculate_torp_ratings()

# View top players
head(ratings)
```

### At a Specific Point in Time

```{r specific}
# Ratings entering round 15 of 2025
ratings <- calculate_torp_ratings(season_val = 2025, round_val = 15)
```

### Game-Level Ratings

For individual game performance (not cumulative):

```{r game}
# All players in round 10
game_ratings <- player_game_ratings(season_val = 2025, round_num = 10)

# Filter by team
carlton <- player_game_ratings(
  season_val = 2025, round_num = 10, team = "Carlton"
)

# Filter by match
match_ratings <- player_game_ratings(matchid = "CD_M20250141001")
```

### Season Totals

```{r season}
# Full season totals
season <- player_season_ratings(2025)

# Partial season
first_half <- player_season_ratings(2025, round_num = 1:12)

# Multiple seasons
multi <- player_season_ratings(2023:2025)
```

## Decay Weighting

TORP uses exponential decay to weight recent games more heavily:

```
weight = exp(-days_since_game / decay_days)
```

The default decay is `RATING_DECAY_DEFAULT_DAYS` (486 days). A game played today has weight 1.0; a game one year ago has weight ~0.47.

```{r decay}
# Shorter decay emphasises recent form
ratings_recent <- calculate_torp_ratings(decay = 200)

# Longer decay for career perspective
ratings_career <- calculate_torp_ratings(decay = 730)
```

## Bayesian Shrinkage

Players with few games are shrunk toward zero to prevent extreme ratings from small samples. Each component has its own prior:

| Component | Prior Games | Effect |
|-----------|------------|--------|
| Receiving | `RATING_PRIOR_GAMES_RECV` (5.87) | ~6 games of zero performance added |
| Disposal  | `RATING_PRIOR_GAMES_DISP` (7.14) | ~7 games of zero performance added |
| Spoil     | `RATING_PRIOR_GAMES_SPOIL` (3) | ~3 games of zero performance added |
| Hitout    | `RATING_PRIOR_GAMES_HITOUT` (3) | ~3 games of zero performance added |

A player with 20 weighted games of disposal data has their rating divided by `20 + 7.14 = 27.14` instead of just `20`, pulling extreme values toward the mean.

## Position Adjustment

Before aggregation, each component's raw credit points are adjusted by subtracting a position-group baseline (the 30th percentile within each position group). This ensures that players are compared fairly across positions -- a midfielder's disposal volume isn't directly compared to a key forward's.

## Pre-Loading Data

For repeated calls (e.g., computing ratings for many rounds), pre-load data once:

```{r preload}
player_game_data <- load_player_game_data(TRUE)
player_details <- load_player_details(2025)

ratings <- calculate_torp_ratings(
  season_val = 2025,
  round_val = 15,
  plyr_tm_df = player_details,
  player_game_data = player_game_data
)
```

## Credit Assignment Parameters

The raw credit points for each component use optimised weights defined in `R/constants.R`. These are tuned via leave-one-season-out cross-validation to minimise RMSE when predicting match margins from team TORP totals. See `data-raw/03-ratings/optimize_torp_ratings.R` for the optimisation procedure.

## See Also

- `vignette("getting-started")` -- Installation and data loading
- `vignette("model-usage")` -- EP, WP, and shot models
- `vignette("season-simulation")` -- Simulating AFL seasons
