---
title: "Season Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Season Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The `simulate_season()` function simulates remaining AFL matches based on team
ratings and fixture data. This is useful for generating probabilistic season
forecasts, finals predictions, and "what if" scenarios.

## Setup

```{r setup}
library(torp)

# Load fixture and results data
fixtures <- load_fixtures(all = TRUE)
results <- load_results(seasons = 2024)
player_stats <- load_player_stats(seasons = 2024)
```

## Preparing Inputs

`simulate_season()` requires two data frames:

1. **sim_teams** — team ratings (one row per team)
2. **sim_games** — fixture data with results for completed games and `NA` for
   upcoming games

```{r prep}
# Calculate team ratings from TORP system
ratings <- torp_ratings(seasons = 2024)

# Build the sim_teams data frame with team-level ratings
sim_teams <- ratings$team_ratings

# Build sim_games from fixtures, marking future games with NA result
sim_games <- fixtures |>
  dplyr::filter(compSeason.year == 2024)
```

## Running a Simulation

```{r run}
# Simulate the season once
sim_result <- simulate_season(sim_teams, sim_games)
```

The function returns a data frame of simulated game results, filling in
scores for games that had `NA` results while preserving actual results for
completed games.

## Multiple Simulations

For probabilistic forecasts, run many simulations and aggregate:

```{r multi}
n_sims <- 1000

all_sims <- lapply(seq_len(n_sims), function(i) {
  sim <- simulate_season(sim_teams, sim_games)
  sim$sim_id <- i
  sim
})

all_sims_df <- data.table::rbindlist(all_sims)
```

## Analysing Results

```{r analyse}
# Count how often each team finishes in each ladder position
library(dplyr)

finals_prob <- all_sims_df |>
  group_by(sim_id) |>
  arrange(desc(wins), desc(percentage)) |>
  mutate(ladder_pos = row_number()) |>
  ungroup() |>
  group_by(team) |>
  summarise(
    top_4_pct = mean(ladder_pos <= 4) * 100,
    top_8_pct = mean(ladder_pos <= 8) * 100,
    avg_wins = mean(wins)
  ) |>
  arrange(desc(top_8_pct))

finals_prob
```

## Key Parameters

The simulation uses constants defined in `R/constants.R`:

- **`SIM_NOISE_SD`** (default: 26) — standard deviation of the random noise
  added to each simulated margin, controlling game-to-game variance
- **`SIM_HOME_ADVANTAGE`** (default: 6) — points added to the home team's
  expected margin
