---
title: "Season Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Season Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The `simulate_season()` function simulates remaining AFL matches based on team ratings and fixture data. Run many simulations to generate probabilistic forecasts for finals qualification, ladder positions, and premiership chances.

## How It Works

For each unplayed match, the simulation:

1. Calculates an expected margin from the team TORP rating difference
2. Adds home ground advantage (`SIM_HOME_ADVANTAGE` = 6 points)
3. Adds random noise (`SIM_NOISE_SD` = 26 points) to reflect game-to-game variance
4. Records the simulated result

Completed matches keep their actual results.

## Running a Simulation

```{r single}
library(torp)

# simulate_season() requires:
# - sim_teams: data frame with team ratings (one row per team)
# - sim_games: fixture data with NA results for unplayed games
sim_result <- simulate_season(sim_teams, sim_games)
```

## Multiple Simulations

For probabilistic forecasts, run many simulations and aggregate:

```{r multi}
n_sims <- 1000

all_sims <- lapply(seq_len(n_sims), function(i) {
  sim <- simulate_season(sim_teams, sim_games)
  sim$sim_id <- i
  sim
})

all_sims_df <- data.table::rbindlist(all_sims)
```

## Analysing Results

```{r analyse}
library(dplyr)

# Finals probability by team
finals_prob <- all_sims_df |>
  group_by(sim_id) |>
  arrange(desc(wins), desc(percentage)) |>
  mutate(ladder_pos = row_number()) |>
  ungroup() |>
  group_by(team) |>
  summarise(
    top_4_pct = mean(ladder_pos <= 4) * 100,
    top_8_pct = mean(ladder_pos <= 8) * 100,
    avg_wins = mean(wins)
  ) |>
  arrange(desc(top_8_pct))

finals_prob
```

## Key Parameters

The simulation uses constants defined in `R/constants.R`:

| Constant | Default | Description |
|----------|---------|-------------|
| `SIM_NOISE_SD` | 26 | Standard deviation of random margin noise |
| `SIM_HOME_ADVANTAGE` | 6 | Points added to home team's expected margin |
| `SIM_WP_SCALING_FACTOR` | 50 | Scaling factor for converting TORP diff to win probability |

## See Also

- `vignette("player-ratings")` -- TORP rating methodology (drives team ratings)
- `vignette("getting-started")` -- Installation and data loading
