---
title: "Player Stat Ratings & PSR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Player Stat Ratings & PSR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

torp's stat rating system estimates each AFL player's "true" per-stat ability using Bayesian models, then combines those estimates into a single **PSR** (Player Skill Rating) that predicts match margin contribution.

The system has two layers:

1. **Stat ratings** -- Bayesian estimates of per-game rates for 48 rate stats (goals, tackles, disposals, ...) and 6 efficiency stats (kick %, disposal %). Uses position-specific priors and exponential time decay so recent form matters more.

2. **PSR** -- A glmnet model that maps stat ratings to predicted margin contribution. PSR tells you how much a player's stat profile is worth in points per game.

```
Raw box-score stats --> Bayesian estimation --> 54 stat ratings per player
                                                      |
                                         glmnet coefficients
                                                      |
                                                     PSR
```

## Loading Pre-Computed Stat Ratings

The easiest way to use stat ratings is to load pre-computed values from torpdata:

```{r load}
library(torp)

# All stat ratings (one row per player-round)
stat_ratings <- load_player_stat_ratings(2026)

# PSR values (predicted margin contribution)
psr <- load_psr(2026)
```

## Player Profiles

View a player's stat rating profile with within-position percentile ranks:

```{r profile}
# Stat rating profile with percentiles (partial name match)
profile <- player_stat_rating_profile("Heeney")
print(profile)

# Full player profile (career stats, TORP ratings, and PSV averages)
full <- player_profile("Heeney")
print(full)
```

`player_stat_rating_profile()` returns a list with:

- `player_info` -- ID, name, team, position group
- `skills` -- Data frame of stat ratings with league-wide and within-position percentile ranks
- `ref_date` -- The reference date used for estimation

## Team Profiles

Aggregate stat ratings to the team level:

```{r team}
# Team stat rating profile
team <- team_stat_rating_profile("Sydney")
print(team)

# Raw team-level stat ratings
team_ratings <- get_team_stat_ratings("Sydney")
```

## Understanding the Numbers

### Stat ratings

Each `*_rating` column represents a Bayesian posterior mean -- the model's best estimate of a player's per-game rate for that stat, adjusted to full TOG (time on ground).

For example, `goals_rating = 2.1` means the model estimates this player would kick 2.1 goals per game at full TOG, given their history, with more weight on recent games.

### PSR, OSR, DSR

| Rating | Full Name | What it measures |
|--------|-----------|-----------------|
| **PSR** | Player Skill Rating | Total predicted margin contribution (points/game above average) |
| **OSR** | Offensive Skill Rating | PSR from attacking stats only |
| **DSR** | Defensive Skill Rating | PSR from defensive stats only |

PSR is centered so that the league-average player scores 0. A PSR of +3.5 means the model predicts this player's stat profile adds 3.5 points per game compared to an average player.

### PSV vs PSR

| | PSV (Value) | PSR (Rating) |
|--|-------------|--------------|
| **Scope** | Single game | Career (with decay) |
| **Input** | Raw game stats | Bayesian-smoothed stat ratings |
| **Use** | "How good was this game?" | "How good is this player?" |

## Computing Stat Ratings From Scratch

If you need custom parameters or a specific reference date:

```{r estimate}
# Load raw player stats
stats <- load_player_stats(TRUE)

# Prepare data for estimation
prep <- .prepare_stat_rating_data(stats)

# Estimate with default parameters
ratings <- estimate_player_stat_ratings(prep, ref_date = as.Date("2026-03-15"))

# Custom hyperparameters
params <- default_stat_rating_params()
params$decay_days <- 200  # faster decay (more recency bias)
params$prior_games <- 8   # weaker prior (less regression to position mean)

ratings_custom <- estimate_player_stat_ratings(prep, params = params)
```

### How the Bayesian Model Works

**Rate stats** (e.g., goals per game) use a Gamma-Poisson conjugate model:

- Prior: Gamma distribution centered on the position-group mean (e.g., forwards' average goal rate)
- Likelihood: Poisson counts from each game, decay-weighted by recency
- Posterior mean: weighted average of prior and observed data

**Efficiency stats** (e.g., kick accuracy %) use a Beta-Binomial model:

- Prior: Beta distribution centered on position-group mean efficiency
- Likelihood: Binomial successes/attempts, decay-weighted
- Posterior mean: smoothed efficiency estimate

The `prior_games` parameter controls how strongly players regress toward position means. Higher values = more regression (useful for players with few games). The `decay_days` parameter controls how quickly old games lose influence.

## Stat Categories

The 48 rate stats span these categories:

| Category | Example Stats |
|----------|--------------|
| Scoring | goals, behinds, shots_at_goal, score_involvements, goal_assists |
| Disposal | kicks, handballs, disposals, effective_kicks, effective_disposals |
| Possession | marks, contested_possessions, uncontested_possessions |
| Contested | contested_marks, ground_ball_gets |
| Clearances | clearances, centre_clearances, stoppage_clearances |
| Inside 50 | inside50s, marks_inside50, rebound50s, metres_gained |
| Defensive | tackles, spoils, intercepts, one_percenters |
| Pressure | pressure_acts, def_half_pressure_acts, tackles_inside50 |
| Ruck | hitouts, hitouts_to_advantage, centre_bounce_attendances, ruck_contests |
| Contests | contest_def_one_on_ones, contest_off_one_on_ones |

The 6 efficiency stats are: kick %, handball %, disposal %, contested possession %, mark %, hitout-to-advantage %.

See `stat_rating_definitions()` for the full list.

## Connecting to TORP

TORP is a 50/50 blend of two independent rating systems:

```
TORP = 50% EPR + 50% PSR
```

- **EPR** comes from play-by-play EPV credit (see `vignette("torp-guide")`)
- **PSR** comes from the stat rating system described here

This separation means EPR captures value from *how* a player uses the ball (context-dependent), while PSR captures *what* a player does statistically (context-independent). The blend provides robustness -- a player who looks good on one system but not the other is rated more cautiously.

## Next Steps

- `vignette("torp-guide")` -- Full reference: EPR, models, data architecture
- `vignette("visualizations")` -- `plot_stat_rating_profile()` and other visualizations
