---
title: "torp Reference Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{torp Reference Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Data Architecture {#data-architecture}

### Package Ecosystem

| Package | Role | Storage |
|---------|------|---------|
| **torp** | Core analytics, data loading, ratings, models | R package |
| **torpdata** | Processed AFL data | GitHub releases (parquet) |
| **torpmodels** | Pre-trained models | GitHub releases (RDS) |

### Data Flow

```
AFL API (get_afl_*) --> torp (clean + enrich) --> torpdata (store)
                            |
                       torpmodels (models)
                            |
                       torp (predict)
```

1. **Collection**: Raw data fetched from the AFL API via torp's `get_afl_*()` functions
2. **Processing**: torp cleans data, adds EP/WP/shot model predictions
3. **Storage**: Processed data uploaded to torpdata GitHub releases as parquet files
4. **Loading**: torp `load_*()` functions download from torpdata on demand
5. **Models**: Trained models stored in torpmodels, loaded via `load_model_with_fallback()`

### Data Types

| Function | Data | Approx Size |
|----------|------|------------|
| `load_pbp()` | Play-by-play events | ~320K rows/season |
| `load_chains()` | Possession sequences | ~160K rows/season |
| `load_player_stats()` | Player game stats | ~10K rows/season |
| `load_player_game_data()` | Player credit points for TORP | ~10K rows/season |
| `load_fixtures()` | Match schedule | ~200 rows/season |
| `load_results()` | Match scores | ~200 rows/season |
| `load_teams()` | Team lineups | ~10K rows/season |
| `load_player_details()` | Player bio info | ~800 rows/season |
| `load_xg()` | Expected goals | ~200 rows/season |
| `load_torp_ratings()` | Pre-computed TORP ratings | ~113K rows total |

### Loading Examples

```{r loading}
library(torp)
pbp <- load_pbp(2025)
pbp <- load_pbp(2025, rounds = 1:10)
pbp_all <- load_pbp(TRUE, rounds = TRUE)
```

### Caching

All data downloads are cached to disk. Fixtures cached in-memory within a session. Models cached in memory to avoid reloading.

```{r caching}
get_disk_cache_dir()
clear_disk_cache()
clear_fixture_cache()
clear_model_cache()
```

### torpdata Releases

| Release Tag | Contents |
|------------|----------|
| `pbp-data` | Play-by-play files by season |
| `chains-data` | Chain files by season |
| `player_stats-data` | Player stat files by season |
| `player_game-data` | Player game credit data by season |
| `fixtures-data` | Fixture files by season |
| `results-data` | Result files by season |
| `teams-data` | Team lineup files by season |
| `player_details-data` | Player detail files by season |
| `xg-data` | Expected goals by season |
| `ratings-data` | Pre-computed TORP ratings |
| `predictions` | Match predictions |

---

## TORP Player Ratings {#player-ratings}

### Overview

TORP (Total Overall Rating Points) evaluates AFL players across four skill categories using play-by-play expected points data.

```
torp = recv/(wt + prior_recv) + disp/(wt + prior_disp)
     + spoil/(wt + prior_spoil) + hitout/(wt + prior_hitout)
```

### Rating Components

**Disposal (torp_disp)**: Value added through kicks and handballs (delta EPV).
**Receiving (torp_recv)**: Value of receiving a disposal (EPV change to receiver).
**Spoil/Tackle (torp_spoil)**: Defensive impact from spoils, tackles, pressure acts.
**Hitout (torp_hitout)**: Ruck contribution from hitouts and ruck contests.

### Getting Ratings

```{r ratings}
library(torp)
ratings <- calculate_torp_ratings()
ratings <- calculate_torp_ratings(season_val = 2025, round_val = 15)

# Game-level
game_ratings <- player_game_ratings(season_val = 2025, round_num = 10)

# Season totals
season <- player_season_ratings(2025)
```

### Decay Weighting

```
weight = exp(-days_since_game / decay_days)
```

Decay varies by component: Receiving 260 days, Spoil 295 days, Disposal/Hitout 700 days. A game one year ago has weight ~0.24 for receiving, ~0.59 for disposal.

### Bayesian Shrinkage

| Component | Prior Games | Decay (days) |
|-----------|------------|--------------|
| Receiving | 12.56 | 260 |
| Disposal  | 5.83 | 700 |
| Spoil     | 3.00 | 295 |
| Hitout    | 15.00 | 700 |

### Pre-Loading Data

```{r preload}
player_game_data <- load_player_game_data(TRUE)
player_details <- load_player_details(2025)
ratings <- calculate_torp_ratings(
  season_val = 2025, round_val = 15,
  plyr_tm_df = player_details,
  player_game_data = player_game_data
)
```

---

## Prediction Models {#models}

Three core models enrich play-by-play data. Models load from [torpmodels](https://github.com/peteowen1/torpmodels) and are cached in memory.

### Complete Pipeline

```{r pipeline}
library(torp)
pbp <- load_pbp(2025, rounds = 1:10)
pbp_clean <- clean_pbp(pbp)
pbp_ep <- add_epv_vars(pbp_clean)        # Expected Points
pbp_wp <- add_wp_vars(pbp_ep)            # Win Probability
shots <- pbp_wp[!is.na(pbp_wp$shot_at_goal), ]
shots <- add_shot_vars(shots)             # Shot Model
```

### Expected Points (EP)

XGBoost model: 5 outcome probabilities. Adds `exp_pts` and `delta_epv`.

### Win Probability (WP)

Ensemble of XGBoost + baseline. Adds `wp`, `wpa`, `wp_category`, `high_leverage`.

### Shot Model

GAM with 3-class probabilities. Adds `goal_prob`, `behind_prob`, `clanger_prob`, `xscore`.

---

## Season Simulation {#simulation}

`simulate_season()` simulates remaining AFL matches using team TORP ratings.

For each unplayed match: expected margin from TORP difference + home advantage (6 pts) + random noise (SD = 26 pts).

```{r sim}
sim_result <- simulate_season(sim_teams, sim_games)

# Multiple simulations
n_sims <- 1000
all_sims <- lapply(seq_len(n_sims), function(i) {
  sim <- simulate_season(sim_teams, sim_games)
  sim$sim_id <- i
  sim
})
all_sims_df <- data.table::rbindlist(all_sims)
```

| Constant | Default | Description |
|----------|---------|-------------|
| `SIM_NOISE_SD` | 26 | Standard deviation of random margin noise |
| `SIM_HOME_ADVANTAGE` | 6 | Points added to home team expected margin |
| `SIM_WP_SCALING_FACTOR` | 50 | Scaling for TORP diff to win probability |