---
title: "How TORP works"
subtitle: "From raw AFL play-by-play to 'who actually added value?'"
author: "Pete Owen"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show the code"
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    embed-resources: true
    smart: false
execute:
  warning: false
  message: false
---

::: {.callout-note}
## The one-paragraph version

**TORP** (Total Over Replacement Predictive-value) is a single number for "how
much did this player add?" It's built from two independent views blended
50/50. **EPV** (Expected Possession Value) reads the actual play-by-play: a
machine-learning model scores every action by how many points it's worth
having the ball *right now*, and the change in that score is credited to
whoever caused it. **PSV** (Player Skill Value) reads the box score instead:
a regression learns which per-stat skills actually predict match margin, then
apportions that margin back to individual players. Win probability (**WP**)
is tracked as a third, parallel view -- it knows the score, EPV deliberately
doesn't -- but it is *not* blended into TORP, for reasons the FAQ at the
bottom explains. Sum a season of credit and you get a rating that says not
just *who* was good, but *how*.
:::

```{mermaid}
%%| fig-cap: "The pipeline at a glance."
flowchart LR
  A["AFL API play-by-play<br/>(daily scrape)"] --> B["Cleaned, standardised PBP<br/>(description whitelist)"]
  B --> C["EP / WP / xG models<br/>(XGBoost + GAM)"]
  C --> D["Credit assignment<br/>(who caused the change?)"]
  D --> E["EPV player ratings"]
  F["Box-score stats"] --> G["Per-stat skill ratings<br/>(Bayesian GAMs)"]
  G --> H["Margin regression<br/>(glmnet -> PSV)"]
  E --> I["TORP = 0.5 x EPV + 0.5 x PSV"]
  H --> I
```

```{r}
#| label: setup
#| output: false
devtools::load_all("../..", quiet = TRUE)
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(ggrepel)
})
theme_set(theme_torp())

# quarto executes chunks with the .qmd's own directory as the working
# directory, so torp's normal getwd()-relative local-data auto-detect
# (get_local_data_dir()) doesn't find the torpdata/ sibling from here --
# point it there explicitly (torp/data-raw/explainer -> torpverse/torpdata/data).
set_local_data_dir(normalizePath("../../../torpdata/data", mustWork = TRUE))

SEASON <- 2026
ROUND  <- 19
MATCH_ID <- "CD_M20260141903"

# One real match, loaded from the local torpdata sibling (zero-network dev):
# Geelong Cats 102 d St Kilda Saints 75, Round 19 2026, GMHBA Stadium.
pbp_round   <- load_pbp(SEASON, rounds = ROUND)
chains_round <- as.data.table(arrow::read_parquet(
  file.path(get_local_data_dir(), "chains_data_2026_all.parquet")
))

pbp_match    <- as.data.table(pbp_round)[match_id == MATCH_ID]
chains_match <- chains_round[match_id == MATCH_ID]
data.table::setorder(pbp_match, display_order)
data.table::setorder(chains_match, display_order)
chains_match[, player_name := paste(player_name_given_name, player_name_surname)]

home_team <- pbp_match$home_team_name[1]
away_team <- pbp_match$away_team_name[1]
final_home <- tail(pbp_match$home_score, 1)
final_away <- tail(pbp_match$away_score, 1)
```

# Step 1 -- The raw data

Everything starts with the **AFL API's play-by-play feed**: a row for every
notable touch of the ball in a match, with a text description, pitch
coordinates and a timestamp. torp's daily scraper pulls this for every match
and publishes it as parquet. In the 2026 season alone that's
`r format(nrow(chains_round), big.mark = ",")` raw rows across
`r uniqueN(chains_round$match_id)` matches -- this document uses one of them,
**`r home_team` `r final_home` d `r away_team` `r final_away`**
(Round `r ROUND`, GMHBA Stadium), which alone contributes
`r nrow(chains_match)` rows.

Here is what the feed actually looks like -- the passage that sets up the
goal this document follows all the way through:

```{r}
#| label: tbl-raw
#| tbl-cap: "Raw play-by-play rows: a turnover deep in defence, intercepted, and worked forward. description is free text; x/y are pitch coordinates."
raw_cols <- c("display_order", "period", "period_seconds", "player_name",
              "description", "x", "y")
raw_window <- chains_match[display_order %between% c(1826, 1841), ..raw_cols]
knitr::kable(raw_window)
```

No sense yet of what any of this is *worth* -- a description string and a
dot on the field. Not something you can model directly. Two things are worth
knowing before any of the field plots later in this document: coordinates
are in real metres from the centre of the ground (this venue is
`r chains_match$venue_length[1]`m x `r chains_match$venue_width[1]`m), and
they are **absolute**, not team-relative -- unlike some sports' event feeds,
the same patch of grass has the same x/y no matter who touches it there.

# Step 2 -- A standard language for actions

Raw description text is noisy -- hundreds of distinct strings, many of them
bookkeeping (subs, timekeeping, video review) rather than an actual passage
of play. torp cleans this down to a whitelist of
`r length(EPV_RELEVANT_DESCRIPTIONS)` play-relevant description types (kicks,
handballs, marks, hard-ball gets, bounces, stoppages...) and engineers a
consistent feature set on top: which quarter, how much time is left, what
kind of play this is, where the ball was a moment ago. This cleaned,
feature-complete table is what the models actually see.

```{r}
#| label: tbl-clean
#| tbl-cap: "The same passage, now modelled. exp_pts is the target the EP model predicts (Step 3): the expected points swing still to come in this quarter, signed from the acting player's team's perspective."
clean_cols <- c("display_order", "player_name", "description", "pos_team",
                "exp_pts", "delta_epv")
clean_window <- pbp_match[display_order %between% c(1832, 1841), ..clean_cols]
knitr::kable(clean_window, digits = c(0, 0, 0, 0, 3, 3))
```

Watch `exp_pts` at the top of that table: `r round(pbp_match[display_order == 1832]$exp_pts, 2)`
on Tobie Travaglia's kick (a St Kilda player, mid-possession) becomes
`r round(pbp_match[display_order == 1833]$exp_pts, 2)` one row later on Jack
Henry's mark -- a **different sign convention is doing the work here**: each
row's `exp_pts` is signed from *that action's own team's* perspective, so the
instant possession changes hands from St Kilda to Geelong, "good news for
St Kilda" flips to "good news for Geelong" even though it's describing
adjacent moments of the same passage. Sit with that sign flip; it's the
entire reason the credit-assignment rules in Step 4 exist.

# Step 3 -- The goal, step by step

Two rows after that intercept, Geelong worked the ball forward in three
touches and scored. This is the single moment this whole document follows
through every remaining stage.

```{r}
#| label: fig-chain
#| fig-cap: "Each numbered label is one action: who did it and what it was. The dotted line is the ball's path; the ground outline is an approximation (torp has no purpose-built ground-drawing helper yet) drawn from this venue's real dimensions."
chain_dt <- pbp_match[chain_number == 231 & period == 4][order(display_order)]
chain_dt[, step := .I]
chain_dt[, clock := sprintf("%d:%02d", period_seconds %/% 60, period_seconds %% 60)]
scorer <- chain_dt[.N, player_name]
assist <- chain_dt[.N - 2, player_name]
goal_clock <- chain_dt[.N, clock]

# Minimal AFL ground outline: an ellipse sized to the real venue dimensions.
# No draw_pitch()/draw_oval() helper exists in torp's plot_*.R yet (checked
# plot_game.R, plot_shots.R) -- plot_shot_map() only draws a half-field with
# reference lines, so this is a deliberately minimal from-scratch outline.
ground_outline <- function(length_m, width_m, n = 200) {
  theta <- seq(0, 2 * pi, length.out = n)
  data.frame(x = (length_m / 2) * cos(theta), y = (width_m / 2) * sin(theta))
}
ground <- ground_outline(chains_match$venue_length[1], chains_match$venue_width[1])

ggplot() +
  geom_path(data = ground, aes(x, y), colour = "grey70", linewidth = 0.6) +
  geom_vline(xintercept = 0, colour = "grey85", linetype = "dashed") +
  geom_path(data = chain_dt, aes(x, y), colour = "grey50",
            linetype = "dotted", linewidth = 0.5,
            arrow = arrow(length = unit(0.15, "cm"), type = "closed")) +
  geom_point(data = chain_dt, aes(x, y), size = 3, colour = "#1B3F8B") +
  ggrepel::geom_text_repel(
    data = chain_dt,
    aes(x, y, label = sprintf("%d. %s\n%s", step, player_name, description)),
    size = 3.2, fontface = "bold", lineheight = 0.95,
    box.padding = 0.7, point.padding = 0.4,
    min.segment.length = 0, seed = 42
  ) +
  coord_fixed() +
  labs(
    title = sprintf("%s's goal, assisted by %s (Q4, %s)", scorer, assist, goal_clock),
    subtitle = sprintf("%s v %s, Round %d %d -- a %d-action passage from turnover to score",
                        home_team, away_team, ROUND, SEASON, nrow(chain_dt)),
    x = NULL, y = NULL
  ) +
  theme_void() +
  theme(plot.title = element_text(face = "bold"), legend.position = "none")
```

The model's job is to put a number on every one of those six actions. Watch
`exp_pts` climb through the passage, then fall straight back to the
post-goal baseline the instant the ball is bounced back to the centre -- the
"cash-in" moment:

```{r}
#| label: fig-epv-timeline
#| fig-cap: "exp_pts through the goal: each action nudges the number up, the goal cashes it in, and the reset after the centre bounce shows just how much of that value was 'used up' by scoring."
timeline_dt <- pbp_match[display_order %between% c(1834, 1843)][order(display_order)]
timeline_dt <- timeline_dt[chain_number == 231 | description == "Centre Bounce"]
timeline_dt[, step := .I]
timeline_dt[, lbl := ifelse(description == "Centre Bounce", "Centre Bounce", player_name)]

ggplot(timeline_dt, aes(step, exp_pts)) +
  geom_step(linewidth = 0.9, colour = "grey40") +
  geom_point(aes(colour = description), size = 3) +
  geom_text(aes(label = lbl), angle = 30, hjust = -0.05, vjust = -0.7, size = 3) +
  scale_x_continuous(breaks = timeline_dt$step, labels = timeline_dt$description,
                      expand = expansion(mult = c(0.05, 0.25))) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.2))) +
  labs(x = NULL, y = "exp_pts (expected points still to come this quarter)") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
```

Zooming out to the whole match: every action, coloured by `exp_pts`. Value
is low in defence, climbs through the midfield, and is highest in each
forward 50 -- exactly where a football fan would expect it, which is the
first sanity check on any model like this.

```{r}
#| label: fig-epv-field
#| fig-cap: "Every action of the match, coloured by exp_pts. Both forward-50s glow -- the model has learned field position without being told anything about football."
ggplot() +
  geom_path(data = ground, aes(x, y), colour = "grey70", linewidth = 0.6) +
  geom_vline(xintercept = 0, colour = "grey85", linetype = "dashed") +
  geom_point(data = pbp_match[!is.na(exp_pts)],
             aes(x, y, colour = pmin(exp_pts, 4)),
             size = 1.4, alpha = 0.65) +
  scale_colour_viridis_c(option = "inferno", name = "exp_pts") +
  coord_fixed() +
  labs(title = sprintf("%s v %s -- where is possession worth something?", home_team, away_team)) +
  theme_void() +
  theme(legend.position = "right")
```

# The model itself

`exp_pts` comes from the **EP (Expected Points) model**: an XGBoost
multiclass classifier that predicts five outcomes for "what's the next score
in this quarter" -- the team in possession kicks a goal or a behind, the
opponent does, or nobody scores again before quarter's end -- and combines
them into a single signed points figure:
`exp_pts = 6 x P(goal) + P(behind) - P(opp behind) - 6 x P(opp goal)`.
It sees 19 features per action:

```{r}
#| label: tbl-features
#| tbl-cap: "The 19 features the EP model sees for every action."
features <- data.frame(
  Feature = c("goal_x, y", "lag_goal_x, lag_goal_x5, lag_y",
              "period_seconds, period", "est_qtr_remaining, est_match_remaining",
              "play_type_kick / handball / reception",
              "phase_of_play_set_shot / hard_ball / loose_ball / handball_received",
              "shot_row", "speed5", "home"),
  `Plain English` = c(
    "Where on the ground the action happens (distance to goal, lateral position)",
    "Where the ball was one and five touches ago",
    "Seconds and quarter number -- which part of the game this is",
    "Time left in the quarter, and in the match",
    "What kind of touch this is",
    "The situation the touch arose from (stoppage, loose ball, general play)",
    "Is this action a shot at goal?",
    "How fast the ball has moved over the last five touches",
    "Is the team in possession the home team?"
  ),
  check.names = FALSE
)
knitr::kable(features)
```

And what the trained model actually leans on:

```{r}
#| label: fig-importance
#| fig-cap: "Feature importance from the live EP model. Time remaining in the quarter dwarfs everything else -- more on why below."
ep_model <- torp:::load_model_with_fallback("ep")
imp <- xgboost::xgb.importance(model = ep_model)
ggplot(as.data.table(imp)[order(Gain)][, Feature := factor(Feature, levels = Feature)],
       aes(Gain, Feature)) +
  geom_col(fill = "#2c7fb8") +
  labs(x = "Share of model gain", y = NULL)
```

`est_qtr_remaining` alone accounts for over half the model's gain -- by far
the single biggest feature, well ahead of field position. That's a genuinely
different story from most possession-value models (which tend to be
dominated by location), and it's worth sitting with rather than glossing
over.

::: {.callout-tip}
## Why "next score in the quarter" and not "final margin"?

Final scores are rare (two teams, one result, once per match) and swamped by
luck late in a blowout. Scores *within a quarter* happen several times per
team per term -- order-of-magnitude more often -- and each one is still a
real, points-denominated outcome. Bounding the target to "this quarter"
rather than "rest of the match" is also what makes the target
*learnable*: a stoppage with 90 seconds left in a quarter genuinely carries
less expected value than the same stoppage with 15 minutes left, because
there is mechanically less time left for a score to happen at all. Training
on the full-match version would blur that real, quarter-shaped signal with
whatever happens after the next break.
:::

::: {.callout-important}
## Isn't `est_qtr_remaining` dominating the model cheating?

It looks suspicious that one clock feature explains over half the model's
predictive power -- shouldn't *football* (field position, what just
happened) be doing more of the work? It isn't cheating, because of what the
target actually is. `exp_pts` is bounded to "before this quarter ends," so
as the clock runs down, the ceiling on how much scoring is even
mechanically possible shrinks toward zero -- regardless of where the ball is.
A model that *didn't* lean heavily on time-remaining would be a worse model,
not a fairer one; it would systematically over-value late-quarter
possessions that have nowhere left to go. What actually isolates one
player's contribution is never the raw level -- it's `delta_epv`, the
*change* from one action to the next. The clock-driven baseline is present
in both the before and after state of every action, so it cancels out of
the difference. This is the same principle as the leakage question in the
football version of this document: what matters is not whether a feature
looks suspicious, but whether it's genuinely settled *before* the thing
being predicted.
:::

# Step 4 -- Giving the value to players

`delta_epv` has to be *earned by someone*. There are four credit rules;
each one below is caught in the act in this actual match, with the row it
applies to marked **&rarr;**.

```{r}
#| label: credit-helper
show_window <- function(rows, cols) {
  win <- pbp_match[display_order %in% rows, ..cols][order(match(display_order, rows))]
  num <- names(win)[vapply(win, is.numeric, logical(1))]
  win[, (num) := lapply(.SD, round, 3), .SDcols = num]
  win
}
```

**Normal disposal + reception** -- no turnover, no contest: the disposer and
the receiver simply split `delta_epv` 50/50 (`EPV_DISP_SCALE` =
`EPV_RECV_SCALE` = `r EPV_DISP_SCALE`). This is the *middle* of our featured
goal: Jack Henry's kick found Gryan Miers on the lead.

```{r}
#| label: tbl-credit-normal
#| tbl-cap: "player_credit = delta_epv x 0.5 for both the kicker and the mark -- an even split because there's no way to say one mattered more than the other."
cols1 <- c("display_order", "player_name", "description", "delta_epv")
win1 <- show_window(c(1835, 1836), cols1)
win1[, role := c("Disposer (kicker)", "Receiver (mark)")]
win1[, credit := round(win1$delta_epv[1] * c(EPV_DISP_SCALE, EPV_RECV_SCALE), 3)]
knitr::kable(win1)
```

**Turnover + intercept mark** -- when the acting player's team *loses* the
ball, torp doesn't split the blame. The player who gave it away eats the
usual 50% disposal share (now strongly negative). But the opponent who reads
the intercept and marks it gets **full** credit for the swing, not half -- 
`EPV_RECV_INTERCEPT_MARK_SCALE` = `r EPV_RECV_INTERCEPT_MARK_SCALE`, not
`r EPV_RECV_SCALE`. Reading the play and taking the mark is treated as
entirely that player's doing. This is the turnover, two rows earlier, that
handed Geelong the ball in the first place:

```{r}
#| label: tbl-credit-turnover
#| tbl-cap: "The kicker's own disposal credit is halved as normal; the intercepting mark gets the FULL swing, not half -- a deliberate asymmetry that rewards reading the play."
win2 <- show_window(c(1832, 1833), cols1)
win2[, role := c("Disposer (turned it over)", "Receiver (intercept mark)")]
win2[, credit := c(round(win2$delta_epv[1] * EPV_DISP_SCALE, 3),
                    round(win2$delta_epv[1] * -1 * EPV_RECV_INTERCEPT_MARK_SCALE, 3))]
knitr::kable(win2)
```

**Three-way aerial contest** -- a kick contested in the air is a genuinely
three-player event (kicker, target, defender), so `compute_contest_credit()`
splits it three ways instead of two: kicker, target and defender each take a
1/3 share (`contest_share`) of the kicker's `delta_epv`, with the target and
defender's shares mirror-signed (one won it, one didn't). Earlier in the
same term, Geelong's James Worpel put one up to Oliver Henry -- this time it
didn't come off:

```{r}
#| label: tbl-credit-contest
#| tbl-cap: "Kicker, target and defender each take a third of the swing. When the target loses the contest (as here), their third is negative and the defender's is positive."
worpel_kick <- pbp_match[display_order == 134]
contest_rows <- data.frame(
  role = c("Kicker (James Worpel)", "Target (Oliver Henry)", "Defender (Jack Silvagni)"),
  credit = round(c(worpel_kick$delta_epv * (1 / 3),
                    worpel_kick$delta_epv * (1 / 3),
                    -worpel_kick$delta_epv * (1 / 3)), 3)
)
knitr::kable(contest_rows)
```

**Box-score credit** -- spoils, tackles, hitouts and a handful of other
defensive/ruck stats never show up as their own row in the modelled
play-by-play at all (`"Spoil"` isn't even in the description whitelist from
Step 2) -- they're credited directly from the official box score, at fixed
weights (`EPV_SPOIL_WT`, `EPV_TACKLE_WT`, `EPV_HITOUT_WT`, ...). Jack
Silvagni again, for his whole game:

```{r}
#| label: tbl-credit-boxscore
#| tbl-cap: "epv_spoil for Jack Silvagni's full match: eight separate defensive box-score stats, each at its own fixed weight, no play-by-play row required."
silvagni <- as.data.table(load_player_stats(SEASON))[match_id == MATCH_ID & grepl("Silvagni", player_name)]
box_tbl <- data.frame(
  stat = c("spoils", "tackles", "pressure_acts", "def_half_pressure_acts",
           "intercepts", "one_percenters", "rebound50s", "frees_against"),
  weight = c(EPV_SPOIL_WT, EPV_TACKLE_WT, EPV_PRESSURE_WT, EPV_DEF_PRESSURE_WT,
             EPV_INTERCEPTS_WT, EPV_ONE_PERCENTERS_WT, EPV_REBOUND50S_WT, EPV_FREES_AGAINST_WT),
  count = as.numeric(silvagni[1, c("spoils", "tackles", "pressure_acts", "def_half_pressure_acts",
                                    "intercepts", "one_percenters", "rebound50s", "frees_against")])
)
box_tbl$contribution <- round(box_tbl$weight * box_tbl$count, 3)
knitr::kable(box_tbl, digits = 4)
cat(sprintf("Total epv_spoil: %.3f points\n", sum(box_tbl$contribution)))
```

Ruck credit works the same way, just off different stats:

```{r}
#| label: tbl-credit-hitout
#| tbl-cap: "epv_hitout for St Kilda's Rowan Marshall, this match: three ruck stats, no play-by-play row required."
marshall <- as.data.table(load_player_stats(SEASON))[match_id == MATCH_ID & grepl("Marshall", player_name)]
hitout_tbl <- data.frame(
  stat = c("hitouts", "hitouts_to_advantage", "ruck_contests"),
  weight = c(EPV_HITOUT_WT, EPV_HITOUT_ADV_WT, EPV_RUCK_CONTEST_WT),
  count = as.numeric(marshall[1, c("hitouts", "hitouts_to_advantage", "ruck_contests")])
)
hitout_tbl$contribution <- round(hitout_tbl$weight * hitout_tbl$count, 3)
knitr::kable(hitout_tbl, digits = 4)
cat(sprintf("Total epv_hitout: %.3f points\n", sum(hitout_tbl$contribution)))
```

Put the whole featured chain back together, and this is what each player
actually banked from it:

```{r}
#| label: tbl-credit-chain
#| tbl-cap: "The featured goal, fully credited. The disposer/receiver split shares value along the chain; the scorer's own kick banks the shot's full delta_epv."
chain_credit <- chain_dt[, .(player_name, description,
                             exp_pts = round(exp_pts, 3),
                             delta_epv = round(delta_epv, 3))]
knitr::kable(chain_credit)
```

# The other half: PSV and player skills

EPV reads the play; **PSV** reads the box score. torp fits ~58 Bayesian
per-stat models (one goals model, one tackles model, one intercepts model...)
that turn each player's raw counting stats into a skill rating, then fits a
single regularised regression (`glmnet`) predicting match margin from a
team's aggregated skill ratings. Each stat's regression coefficient
(`beta`) says how much that skill actually moves the scoreboard; apportioning
those coefficients back to individual players is PSV -- "predicted margin
contribution, in points, above a replacement player."

The coefficients currently shipped were fit on 2024 data. A 2026 research
pass refit them on 2025 to check for drift -- which stats the margin model
leans on *did* move, meaningfully, for some stats:

```{r}
#| label: fig-beta-drift
#| fig-cap: "The 15 stats whose PSR coefficient moved most between the 2024 and 2025 refits. Positive delta = the stat became more heavily rewarded in the newer fit."
beta_path <- "../../../torpmodels/data-raw/04-match-model/experiments/results/ws1_beta_drift.csv"
beta_drift <- as.data.table(read.csv(beta_path))
top_drift <- beta_drift[order(-abs_delta)][1:15]
top_drift[, stat_name := factor(stat_name, levels = stat_name[order(delta)])]
ggplot(top_drift, aes(delta, stat_name, fill = delta > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c(`TRUE` = "#1B9E77", `FALSE` = "#D95F02")) +
  labs(x = "beta_v2025 - beta_v2024", y = NULL,
       title = "Which stats moved most in a 2025 PSR refit")
```

The biggest mover is `intercepts` -- its 2024 coefficient (6.77) roughly
halved in the 2025 refit -- while stats like `metres_gained`, `behinds` and
`marks_inside50` went from *zero* weight (excluded by the regulariser
entirely) to real, positive coefficients. This was a real, measured finding
from a 2026 research pass (`WS1`) -- but per that investigation's own
conclusion, the drift isn't costing measurable match-prediction accuracy
this season, so the shipped coefficients stayed on the 2024 vintage rather
than triggering an annual-refresh policy. It's included here because "which
stats does the model value, and is that stable" is exactly the kind of
question this document exists to make answerable.

# Win probability: a parallel view

Every action also gets a **WP** (win probability) and **WPA** (win
probability added) figure, from a separate GAM/XGBoost model that -- unlike
EPV -- is allowed to see the score. WP is served **calibrated**: a
leverage-interaction sidecar fitted on temporal holdout data, which brought
the model's late-game/close-game slope from roughly 1.75 down to about 0.95
(i.e. corrected a real overconfidence problem in tight finishes).

```{r}
#| label: fig-wp
#| fig-cap: "Win probability across the whole match. Home = Geelong Cats."
plot_ep_wp(SEASON, ROUND, MATCH_ID, metric = "wp")
```

WPA is credited the same disposer/receiver way as EPV
(`WP_CREDIT_DISP_SHARE` = `r WP_CREDIT_DISP_SHARE`), but it is tracked
**only as a parallel display metric** -- it is not one of the two halves that
make up `torp_value`. The FAQ below explains exactly why, because the
reasoning changed in 2026 and it's a better story than it used to be.

# Step 5 -- A season of credit: player value profiles

Sum a season of EPV credit per player, split by channel, normalise per game
 -- and two players who both rate well can look completely different. Here's
an actual 2026 archetype contrast: **Max Gawn**, a ruck (2nd in the league
by hitout credit), against **Charlie Cameron**, a small forward (top-5 by
total EPV among medium forwards).

```{r}
#| label: fig-profiles
#| fig-cap: "Two very different ways to add value. Gawn's value is concentrated in hitouts (a channel Cameron barely touches); Cameron's comes almost entirely from receiving and disposing the ball inside attacking 50."
pgd_season <- create_player_game_data(
  pbp_data = load_pbp(SEASON, rounds = TRUE),
  player_stats = as.data.table(load_player_stats(SEASON)),
  teams = as.data.table(load_teams(SEASON)),
  chains = as.data.table(arrow::read_parquet(
    file.path(get_local_data_dir(), "chains_data_2026_all.parquet")
  ))
)
pgd_season <- as.data.table(pgd_season)

two_players <- c("Max Gawn", "Charlie Cameron")
season_totals <- pgd_season[player_name %in% two_players,
  .(games = .N, recv = sum(epv_recv), disp = sum(epv_disp),
    spoil = sum(epv_spoil), hitout = sum(epv_hitout)),
  by = player_name]
long <- data.table::melt(season_totals, id.vars = c("player_name", "games"),
                         variable.name = "area", value.name = "credit")
long[, credit_pg := credit / games]

# shared area order + shared x scale across both facets, so the two profiles
# read across for comparison
area_order <- long[, .(total = sum(credit_pg)), by = area][order(total), area]
long[, area := factor(area, levels = area_order)]

ggplot(long, aes(credit_pg, area, fill = player_name)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~player_name) +
  scale_fill_manual(values = c("Max Gawn" = team_color_lookup("Melbourne Demons"),
                                "Charlie Cameron" = team_color_lookup("Brisbane Lions"))) +
  labs(x = "EPV credit per game (points)", y = NULL,
       title = sprintf("Where does each player's value come from? (%d season)", SEASON),
       subtitle = "Same four channels, same order, same scale -- read across to compare")
```

All four bars are **EPV credit channels**, not raw action types -- `recv`
and `disp` are the disposer/receiver split from Step 4, `spoil` is the
box-score defensive channel, `hitout` is the ruck channel. The units
throughout are **points**: Gawn's ~`r round(season_totals[player_name == "Max Gawn"]$hitout / season_totals[player_name == "Max Gawn"]$games, 2)`
points of hitout credit per game is a real, if modest, points swing his
ruck work is worth on average, every game, on top of everything else he
does with ball in hand.

# The questions you should be asking

Two of the best ones -- *why next-score-in-quarter instead of final margin*,
and *isn't `est_qtr_remaining` dominating cheating* -- are answered up in the
model section. Here are the rest.

::: {.callout-note collapse="true"}
## How do you know the model is any good?

Three layers, the same as any serious rating system. First, standard ML
hygiene: EP and WP are trained and release-gated on out-of-sample data -- 
WP specifically has to pass a temporal-holdout slope gate
(`|calibrated slope - 1| <= 0.10`, both overall and in the Q4/close-game
cell specifically) before a retrain is even allowed to publish. Second,
sanity: the feature-importance chart and the field heatmap both reproduce
football intuition (forward-50 is valuable, time matters) without being
told anything about the sport. Third, and the one that matters most: the
same EPV/PSV ratings feed a match-prediction model that is scored against
real results every single week. A rating system that wasn't measuring real
value would wash out there.
:::

::: {.callout-note collapse="true"}
## Does the model know the score?

Depends which model. **EP is deliberately score-blind** -- look back at the
feature list, there's a clock but no `points_diff`. A team already up big
stops attacking normal-game-state footy, and if EP conditioned on score,
every player's actions in a blowout would be re-valued by context they
don't control. **WP is the opposite on purpose** -- it exists specifically
to answer "given the score and the clock, who's going to win," so
`points_diff` and friends are core features. That's the real reason WPA
stays out of `torp_value` even though it's a perfectly good model in its
own right -- see the next question.
:::

::: {.callout-important collapse="true"}
## Why isn't WPA blended into TORP, if it's already built?

This used to be answered with a claim that turned out to be false: "the WP
gradient is too steep late/close, so blending it in would distort things."
Measured directly in mid-2026, that wasn't true -- the calibrated WP family
is actually *flat* in exactly that region. So the door to reinstating WPA
was reopened, with a pre-registered bar: if a better-calibrated WP model
*shrank* the tendency for WPA to reward close-game exposure by more than
50%, blending it back in would become defensible.

It didn't clear that bar -- and the direction of the miss is the
interesting part. Calibration *dropped the correlation* between a player's
WPA rate and how much of their time on ground came in close games
(0.164 to 0.116), but the *regression coefficient* on that exposure -- how
big the effect actually is per unit of exposure -- went the other way,
**up** 47%, from 0.0284 to 0.0418. In other words: a sharper, better-behaved
WP model didn't shrink the close-game-exposure pattern in WPA, it sharpened
it. That's the tell that this isn't a calibration artifact at all -- it's
structural. WPA, by construction, measures more when the game is live and
close; that's not a bug to be calibrated away, it's what the metric *is*.
Blending it into a season-long, context-neutral rating like TORP would mean
systematically rewarding players who happened to get more high-leverage
minutes, which is closer to noise than skill. WPA stays exactly where it
is: a genuinely useful parallel display metric, not a TORP ingredient -- and
the decision now rests on a measurement, not the disproven gradient story.
:::

::: {.callout-note collapse="true"}
## Couldn't a player farm value by just having a shot from anywhere?

No -- a shot's value is judged against the situation it came from, the same
as any other action. A speculative snap from the boundary carries a low
`exp_pts` of its own; missing it produces a strongly negative `delta_epv`
that the shooter banks in full (Step 4's "normal disposal" rule -- nobody
shares the blame for a bad shot). The shooting leaderboard rewards
**selection and finishing position**, not volume: a player who shoots less
often but only from good position out-earns a high-volume speculator every
time.
:::

::: {.callout-note collapse="true"}
## Why XGBoost, and why glmnet for PSR?

Different jobs, different tools. EP/WP/xG are tabular, few-hundred-thousand
row problems where gradient-boosted trees are still the thing to beat -- 
fast to retrain, no GPU needed, and the feature-importance chart above is a
free byproduct. PSR is a different shape of problem: ~50 correlated skill
ratings predicting one number (match margin) per team per round, where the
goal is stable, interpretable *coefficients* rather than raw predictive
power -- that's exactly what a regularised linear model (`glmnet`) is built
for, and it's also why the WS1 beta-drift comparison above is even a
meaningful thing to plot.
:::

::: {.callout-note collapse="true"}
## What can event data not see at all?

Anything off the ball. The play-by-play records who touched it and where -- 
it has no idea about a defender's positioning that made a pass impossible
to attempt, a lead that dragged two opponents out of a contest, or a
tackler running down a play 30 metres away that never shows up in a stat
line. The box-score-driven PSV channel closes some of this gap (pressure
acts, one-percenters, defensive half pressure are all rewarded even without
a corresponding play-by-play row), but torp doesn't currently have a
plus-minus layer analogous to panna's RAPM to catch what's left. That's a
real, acknowledged limitation, not a hidden one.
:::

# Where this goes next

EPV and PSV are already blended, live, every day -- `torp_value = 0.5 x epv +
0.5 x psv` (`TORP_EPR_WEIGHT`). Downstream of that:

- **Match predictions** use team-aggregated TORP/EPR/PSR differentials plus a
  sequential team-Elo feature, recalibrated against real results every
  retrain;
- **WPA** keeps improving in its own right (the calibration sidecar above
  shipped in July 2026) even while it stays out of the blend;
- Everything here is republished automatically, every day, from the same
  pipeline this document just walked through.

::: {.callout-note collapse="true"}
## Reproducing this document

Everything here runs from the torp repo against the local torpdata sibling
(`../../../torpdata/data` from this file, zero network calls beyond the model
download and one live box-score refresh) -- no committed fixtures, just the same daily
data the production pipeline uses. Render with
`quarto render data-raw/explainer/torp_explainer.qmd` from the `torp/` root
(via PowerShell -- arrow segfaults under Git Bash R on Windows).
:::
